Efficient pedigree recording for fast population genetics simulation

### 2018-01-11 forwards_paper.tex
Move 'next steps' to discussion

\paragraph{Validity of a set of tables}
Given a set of node and edge tables as described above,
there are only two requirements that ensure the tables
describe a valid tree sequence.
% (There are essentially no such requirements on the site and mutation tables.)
These are:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
A pair of node and edge tables that satisfy these two requirements
is guaranteed to uniquely describe at each point on the genome
a collection of directed, acyclic graphs -- in other words, a forest of trees.
For some applications it is necessary to check that at every point
there is only a \emph{single} tree.
Checking this is more difficult, but is implemented in \msprime{}'s API.
% \jda{TODO add function name}  PLR: not sure this is necessary.
For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
\msprime{}'s API includes tools to rectify this by first sorting % (using \texttt{sort\_tables})
and then using the \texttt{simplify} algorithm described below, which works on sorted tables
and is guaranteed to produce a valid, \msprime{}-ready tree sequence.

@@

\paragraph{Flexibility.}
To demonstrate the flexibility provided by the Tables API and provide an
implementation that decouples forward simulation internals from transfer of data
to \msprime\, we implemented a version of the simulations described in
``Simulation benchmarks'' separately in Python (see Appendix~\ref{ss:simupop}).
For this proof-of-concept implementation, the simulation engine invokes
callbacks at critical points of the simulation, and we infer nodes and edges
from the information that is provided. Rows are appended to the tables
one-by-one, and the tables are periodically sorted and simplified to control
memory usage.
It uses the \simupop{} library, interfacing with Python code (via our
package \ftprime{}) to simulate a discrete but large number of sites instead of the
infinite sites model; we provide results from this implementation alongside the
results from the \cpp implementation for simulations without selection  (see
Appendix~\ref{ss:timing_nosel}).
% $\mu$ and $r$ define per-base-pair mutation rate a
% the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.  We
% varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.

\paragraph{Next steps.}
While these methods work well, and indeed result in significant performance
gains over non-ancestry based simulation, they are far from optimal. Because
forwards time simulation is highly CPU intensive, most simulation engines
are written in C or C++ out of necessity. Therefore, the simplest and most
efficient approach would be to call the \msprime\ C library directly as part of
the core simulation logic, in the manner outlined in Algorithm W. If designed
in this way, the simulation engine would not need to keep copies of the node
and edge information (and in the more flexible \simupop{} implementation, would
not need to infer edges);
even greater gains in computational efficiency could then be obtained. To
facilitate this usage, a standalone, embeddable C library encompassing the core
tree sequence functionality is planned, which should make recording and
outputting tree sequences from any C/C++ code straightforward.

@@

A final note:
in preparing this manuscript,
we debated a number of possible terms for the embellished pedigree,
i.e., the ``pedigree with ancestral recombination information''.
Etymological consensus \citep{liberman2014little} has
``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
(whose branching pattern resembles the bifurcation of a single parent-offspring relationship).
An analogous term for the embellished pedigree might then be \emph{nedigree},
from ``nid de grue'',
as the nest of a crane is a large jumble of (forking) branches.
We thought it unwise to use this term throughout the manuscript,
but perhaps it will prove useful elsewhere.

>>>>

\paragraph{Validity of a set of tables}
Given a set of node and edge tables as described above,
there are only two requirements that ensure the tables
describe a valid tree sequence.
% (There are essentially no such requirements on the site and mutation tables.)
These are:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
A pair of node and edge tables that satisfy these two requirements
is guaranteed to uniquely describe at each point on the genome
a collection of directed, acyclic graphs -- in other words, a forest of trees.
For some applications it is necessary to check that at every point
there is only a \emph{single} tree.
Checking this is more difficult, but is implemented in \msprime{}'s API.
For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
\msprime{}'s API includes tools to rectify this by first sorting % (using \texttt{sort\_tables})
and then using the \texttt{simplify} algorithm described below, which works on sorted tables
and is guaranteed to produce a valid, \msprime{}-ready tree sequence.

@@

\paragraph{Flexibility.}
To demonstrate the flexibility provided by the Tables API and provide an
implementation that decouples forward simulation internals from transfer of data
to \msprime\, we implemented a version of the simulations described in
``Simulation benchmarks'' separately in Python (see Appendix~\ref{ss:simupop}).
For this proof-of-concept implementation, the simulation engine invokes
callbacks at critical points of the simulation, and we infer nodes and edges
from the information that is provided. Rows are appended to the tables
one-by-one, and the tables are periodically sorted and simplified to control
memory usage.
It uses the \simupop{} library, interfacing with Python code (via our
package \ftprime{}) to simulate a discrete but large number of sites instead of the
infinite sites model; we provide results from this implementation alongside the
results from the \cpp\ implementation for simulations without selection  (see
Appendix~\ref{ss:timing_nosel}).
% $\mu$ and $r$ define per-base-pair mutation rate a
% the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.  We
% varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.

@@

\paragraph{Next steps.}
While our methods work well, and indeed result in significant performance
gains over non-ancestry based simulation, they are far from optimal. Because
forwards time simulation is highly CPU intensive, most simulation engines
are written in C or C++ out of necessity. Therefore, the simplest and most
efficient approach would be to call the \msprime\ C library directly as part of
the core simulation logic, in the manner outlined in Algorithm W. If designed
in this way, the simulation engine would not need to keep copies of the node
and edge information (and in the more flexible \simupop{} implementation, would
not need to infer edges);
even greater gains in computational efficiency could then be obtained. To
facilitate this usage, a standalone, embeddable C library encompassing the core
tree sequence functionality is planned, which should make recording and
outputting tree sequences from any C/C++ code straightforward.

\paragraph{A final note:}
in preparing this manuscript,
we debated a number of possible terms for the embellished pedigree,
i.e., the ``pedigree with ancestral recombination information''.
Etymological consensus \citep{liberman2014little} has
``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
(whose branching pattern resembles the bifurcation of a single parent-offspring relationship).
An analogous term for the embellished pedigree might then be \emph{nedigree},
from ``nid de grue'',
as the nest of a crane is a large jumble of (forking) branches.
We thought it unwise to use this term throughout the manuscript,
but perhaps it will prove useful elsewhere.
